---
title: "Homework 2"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
```{r setup, message = F, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(DAAG)
```

The data set calif_penn_2011.csv contains information about the housing stock of California and Pennsylvania, as of 2011.  Information as aggregated into "Census tracts", geographic regions of a few thousand people which are supposed to be fairly homogeneous economically and socially.

1. _Loading and cleaning_
    a. Load the data into a dataframe called `ca_pa`.
    b. How many rows and columns does the dataframe have?
    c. Run this command, and explain, in words, what this does:
```
colSums(apply(ca_pa,c(1,2),is.na))
```
    d. The function `na.omit()` takes a dataframe and returns a new dataframe, omitting any row containing an NA value.  Use it to purge the data set of rows with incomplete data.
    e. How many rows did this eliminate?
    f. Are your answers in (c) and (e) compatible? Explain.
```{r}
ca_pa <- read.csv("data/calif_penn_2011.csv", sep = ',', header=T)

dim_info <- dim(ca_pa)
paste("Rows:", dim_info[1], "Columns:", dim_info[2])

colSums(apply(ca_pa,c(1,2),is.na))

ca_pa_clean <- na.omit(ca_pa)

rows_eliminated <- nrow(ca_pa) - nrow(ca_pa_clean)
paste("Rows eliminated:", rows_eliminated)
```
#c
This command calculates the number of missing values (NA) in each column of the dataframe.
apply(ca_pa, c(1,2), is.na) converts each element to TRUE if it's NA, FALSE otherwise
colSums() then sums the TRUE values (NA count) column-wise
#f
Yes, they are compatible. The value in (e) should equal the number of rows with at least one NA (missing value), while the sum of values in (c) counts all missing values across all cells. Since a row may have multiple NAs, the sum in (c) will typically be larger than (e), but both reflect the presence of missing data.

2. _This Very New House_  
    a. The variable `Built_2005_or_later` indicates the percentage of houses in each Census tract built since 2005.  Plot median house prices against this variable.
    b. Make a new plot, or pair of plots, which breaks this out by state.  Note that the state is recorded in the `STATEFP` variable, with California being state 6 and Pennsylvania state 42.
```{r}
ggplot(ca_pa_clean, aes(x = Built_2005_or_later, y = Median_house_value)) +
  geom_point(alpha = 0.6) +
  labs(x = "Percentage built since 2005", y = "Median house value",
       title = "House Value vs New Houses")

ggplot(ca_pa_clean, aes(x = Built_2005_or_later, y = Median_house_value, color = factor(STATEFP))) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ STATEFP, labeller = labeller(STATEFP = c("6" = "California", "42" = "Pennsylvania"))) +
  labs(x = "Percentage built since 2005", y = "Median house value",
       title = "House Value vs New Houses by State", color = "State")
```
3. _Nobody Home_  
The vacancy rate is the fraction of housing units which are not occupied.  The dataframe contains columns giving the total number of housing units for each Census tract, and the number of vacant housing units.  
    a. Add a new column to the dataframe which contains the vacancy rate.  What are the minimum, maximum, mean, and median vacancy rates?
    b. Plot the vacancy rate against median house value.
    c. Plot vacancy rate against median house value separately for California and for Pennsylvania.  Is there a difference?
    
```{r}
ca_pa_clean <- ca_pa_clean %>%
  mutate(vacancy_rate = (Vacant_units / Total_units) * 100)

vacancy_stats <- ca_pa_clean %>%
  summarize(
    min = min(vacancy_rate),
    max = max(vacancy_rate),
    mean = mean(vacancy_rate),
    median = median(vacancy_rate)
  )
vacancy_stats

ggplot(ca_pa_clean, aes(x = vacancy_rate, y = Median_house_value)) +
  geom_point(alpha = 0.5) +
  labs(x = "Vacancy rate (%)", y = "Median house value",
       title = "House Value vs Vacancy Rate")

ggplot(ca_pa_clean, aes(x = vacancy_rate, y = Median_house_value)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ STATEFP, labeller = labeller(STATEFP = c("6" = "California", "42" = "Pennsylvania"))) +
  labs(x = "Vacancy rate (%)", y = "Median house value",
       title = "House Value vs Vacancy Rate by State")
```
 There appears to be a difference between states. California shows a tighter cluster of high-value properties with low vacancy rates, while Pennsylvania has more variability and generally lower house values.
    
4. The column `COUNTYFP` contains a numerical code for counties within each state.  We are interested in Alameda County (county 1 in California), Santa Clara (county 85 in California), and Allegheny County (county 3 in Pennsylvania).
    a. Explain what the block of code at the end of this question is supposed to accomplish, and how it does it.
The code calculates the median house value (column 10) for Alameda County (CA) by:
Looping through all tracts to identify those in California (STATEFP=6) and Alameda County (COUNTYFP=1)
Storing qualifying tract indices in acca
Extracting Median_house_value for these tracts into accamhv
Calculating the median of these values
    
    b. Give a single line of R which gives the same final answer as the block of code.  Note: there are at least two ways to do this; you just have to find one.
```{r}    
    median(ca_pa_clean$Median_house_value[ca_pa_clean$STATEFP == 6 & ca_pa_clean$COUNTYFP == 1])
```
    c. For Alameda, Santa Clara and Allegheny Counties, what were the average percentages of housing built since 2005?
```{r}    
ca_pa_clean %>%
  filter(
    (STATEFP == 6 & COUNTYFP == 1) |  # Alameda
    (STATEFP == 6 & COUNTYFP == 85) | # Santa Clara
    (STATEFP == 42 & COUNTYFP == 3)   # Allegheny
  ) %>%
  group_by(COUNTYFP, STATEFP) %>%
  summarize(avg_new_housing = mean(Built_2005_or_later))
```
    d. The `cor` function calculates the correlation coefficient between two variables.  What is the correlation between median house value and the percent of housing built since 2005 in (i) the whole data, (ii) all of California, (iii) all of Pennsylvania, (iv) Alameda County, (v) Santa Clara County and (vi) Allegheny County?
```{r}    
# Whole data
cor(ca_pa_clean$Median_house_value, ca_pa_clean$Built_2005_or_later)

# California
ca_data <- filter(ca_pa_clean, STATEFP == 6)
cor(ca_data$Median_house_value, ca_data$Built_2005_or_later)

# Pennsylvania
pa_data <- filter(ca_pa_clean, STATEFP == 42)
cor(pa_data$Median_house_value, pa_data$Built_2005_or_later)

# Alameda
alameda <- filter(ca_pa_clean, STATEFP == 6, COUNTYFP == 1)
cor(alameda$Median_house_value, alameda$Built_2005_or_later)

# Santa Clara
santa_clara <- filter(ca_pa_clean, STATEFP == 6, COUNTYFP == 85)
cor(santa_clara$Median_house_value, santa_clara$Built_2005_or_later)

# Allegheny
allegheny <- filter(ca_pa_clean, STATEFP == 42, COUNTYFP == 3)
cor(allegheny$Median_house_value, allegheny$Built_2005_or_later)
```
    e. Make three plots, showing median house values against median income, for Alameda, Santa Clara, and Allegheny Counties.  (If you can fit the information into one plot, clearly distinguishing the three counties, that's OK too.)
```{r}    
county_data <- ca_pa_clean %>%
  filter(
    (STATEFP == 6 & COUNTYFP == 1) |  # Alameda
    (STATEFP == 6 & COUNTYFP == 85) | # Santa Clara
    (STATEFP == 42 & COUNTYFP == 3)   # Allegheny
  ) %>%
  mutate(county = case_when(
    COUNTYFP == 1 ~ "Alameda",
    COUNTYFP == 85 ~ "Santa Clara",
    COUNTYFP == 3 ~ "Allegheny"
  ))

ggplot(county_data, aes(x = Median_household_income, y = Median_house_value, color = county)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ county, scales = "free") +
  labs(x = "Median Income", y = "Median House Value",
       title = "House Value vs Income by County")
```    

```{r, eval=FALSE}
acca <- c()
for (tract in 1:nrow(ca_pa)) {
  if (ca_pa$STATEFP[tract] == 6) {
    if (ca_pa$COUNTYFP[tract] == 1) {
      acca <- c(acca, tract)
    }
  }
}
accamhv <- c()
for (tract in acca) {
  accamhv <- c(accamhv, ca_pa[tract,10])
}
median(accamhv)
```

MB.Ch1.11. Run the following code:
```{r}
gender <- factor(c(rep("female", 91), rep("male", 92)))
table(gender)
gender <- factor(gender, levels=c("male", "female"))
table(gender)
gender <- factor(gender, levels=c("Male", "female"))
table(gender)
table(gender, exclude=NULL)
rm(gender) 
```
Explain the output from the successive uses of table().
First table(): Counts factor levels as created (91 female, 92 male)
Second table(): Same counts but with level order reversed (male first)
Third table(): Shows 0 for "Male" because original "male" doesn't match new level "Male" (case mismatch)
Fourth table(exclude=NULL): Includes NA counts (92) from mismatched "male" values

MB.Ch1.12. Write a function that calculates the proportion of values in a vector x that exceed some value cutoff.

(a) Use the sequence of numbers 1, 2, . . . , 100 to check that this function gives the result that is expected.
```{r}
prop_above <- function(x, cutoff) {
  mean(x > cutoff, na.rm = TRUE)
}
prop_above(1:100, 50) 
prop_above(1:100, 90) 
```

(b) Obtain the vector ex01.36 from the Devore6 (or Devore7) package. These data give the times required for individuals to escape from an oil platform during a drill. Use dotplot() to show the distribution of times. Calculate the proportion of escape times that exceed 7 minutes.
```{r}
library(Devore7)
data(ex01.36)
prop_above <- function(x, threshold) {
  mean(x > threshold)
}
library(lattice)
dotplot(~ex01.36, xlab = "Escape Time (s)", main = "proportion of escape times")
prop_above(ex01.36,420)
```

MB.Ch1.18. The Rabbit data frame in the MASS library contains blood pressure change measurements on five rabbits (labeled as R1, R2, . . . ,R5) under various control and treatment conditions. Read the help file for more information. Use the unstack() function (three times) to convert Rabbit to the following form:

Treatment Dose   R1   R2   R3   R4   R5 

1 Control 6.25  0.50 1.00 0.75 1.25 1.5 

2 Control 12.50 4.50 1.25 3.00 1.50 1.5 

....
```{r}
 library(MASS)
data(Rabbit)

Rabbit <- Rabbit %>%
  mutate(group = interaction(Treatment, Dose))

Rabbit_wide <- do.call(cbind, lapply(split(Rabbit, Rabbit$Animal), 
                                     function(df) unstack(df, BPchange ~ group)))

Rabbit_final <- data.frame(
  Treatment = sapply(strsplit(rownames(Rabbit_wide), "\\."), `[`, 1),
  Dose = as.numeric(sapply(strsplit(rownames(Rabbit_wide), "\\."), `[`, 2)),
  Rabbit_wide
) %>% `rownames<-`(NULL)

head(Rabbit_final, 4)
```

