---
title: "Homework 1"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
```{r setup, message = F, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(DAAG)
```

1. The Iowa data set `iowa.csv` is a toy example that summarises the yield of wheat (bushels per acre) for the state of Iowa between 1930-1962. In addition to yield, year, rainfall and temperature were recorded as the main predictors of yield.
    a. First, we need to load the data set into R using the command `read.csv()`. Use the help function to learn what arguments this function takes. Once you have the necessary input, load the data set into R and make it a data frame called `iowa.df`.
    b. How many rows and columns does `iowa.df` have? 
    c. What are the names of the columns of `iowa.df`?
    d. What is the value of row 5, column 7 of `iowa.df`?
    e. Display the second row of `iowa.df` in its entirety.
```{r}
iowa.df<-read.csv("data/Iowa.csv", sep = ',', header=T)
# b. 
cat("b. Number of rows:", nrow(iowa.df), "\n")
cat("   Number of columns:", ncol(iowa.df), "\n\n")
# c.
cat("c. Column names:", paste(names(iowa.df), collapse = ", "), "\n\n")
# d. 
cat("d. Value at row 5, column 7:", iowa.df[5,7], "\n\n")
# e. 
cat("e. Second row:\n")
print(iowa.df[2, ])
```

2. Syntax and class-typing.
    a. For each of the following commands, either explain why they should be errors, or explain the non-erroneous result.

```
vector1 <- c("5", "12", "7", "32")
[1] "5"  "12" "7"  "32"
max(vector1)
[1] "7"
sort(vector1)
[1] "12" "32" "5"  "7" 
sum(vector1)
Error: invalid 'type' (character) of argument
sum() requires numeric input
```
    b. For the next series of commands, either explain their results, or why they should produce errors.

```
vector2 <- c("5",7,12)
vector2[2] + vector2[3]
Error: cannot add strings
When creating the vector vector2, since there is a string "5", R will coerce the entire vector to the character type. When executing vector2[2] + vector2[3], it actually attempts to perform addition on the character values "7" and "12". Since character types cannot be directly used in arithmetic operations, an error will occur.

dataframe3 <- data.frame(z1="5",z2=7,z3=12)
dataframe3[1,2] + dataframe3[1,3]
 7 + 12 = 19
 
list4 <- list(z1="6", z2=42, z3="49", z4=126)
list4[[2]]+list4[[4]]
42 + 126 = 168
list4[2]+list4[4]
Error: cannot add sublists
list4[2] and list4[4] return sublists, not individual elements
```

3. Working with functions and operators.
    a. The colon operator will create a sequence of integers in order. It is a special case of the function `seq()` which you saw earlier in this assignment. Using the help command `?seq` to learn about the function, design an expression that will give you the sequence of numbers from 1 to 10000 in increments of 372. Design another that will give you a sequence between 1 and 10000 that is exactly 50 numbers in length.
```{r}
cat("Results:\n")
seq_by_372 <- seq(1, 10000, by = 372)
cat("Sequence by 372 (first 5 values):", head(seq_by_372, 5), "...\n")

seq_length_50 <- seq(1, 10000, length.out = 50)
cat("Sequence with 50 elements (first/last):", 
    head(seq_length_50, 2), "...", tail(seq_length_50, 2), "\n\n")
```

    b. The function `rep()` repeats a vector some number of times. Explain the difference between `rep(1:3, times=3)` and `rep(1:3, each=3)`.
```{r}
cat("Results:\n")
cat("rep(1:3, times=3):", rep(1:3, times = 3), "\n")
cat("rep(1:3, each=3):", rep(1:3, each = 3), "\n")
cat("Difference: 'times' repeats whole vector, 'each' repeats each element\n\n")
```
MB.Ch1.2. The orings data frame gives data on the damage that had occurred in US space shuttle launches prior to the disastrous Challenger launch of 28 January 1986. The observations in rows 1, 2, 4, 11, 13, and 18 were included in the pre-launch charts used in deciding whether to proceed with the launch, while remaining rows were omitted.

Create a new data frame by extracting these rows from orings, and plot total incidents against temperature for this new data frame. Obtain a similar plot for the full data set.
```{r}
data(orings)
selected_rows <- c(1, 2, 4, 11, 13, 18)
orings_sub <- orings[selected_rows, ]

plot(Total ~ Temperature, data = orings_sub,
     main = "O-Ring Damage (Pre-Challenger Subset)",
     xlab = "Temperature (°F)", ylab = "Damage Incidents",
     pch = 19, col = "red")

plot(Total ~ Temperature, data = orings,
     main = "O-Ring Damage (Complete Dataset)",
     xlab = "Temperature (°F)", ylab = "Damage Incidents")
points(orings_sub$Temperature, orings_sub$Total, 
       pch = 19, col = "red")
legend("topright", legend = c("Used in decision", "Omitted"),
       pch = c(19, 1), col = c("red", "black"))
```

MB.Ch1.4. For the data frame `ais` (DAAG package)
data(ais)
(a) Use the function `str()` to get information on each of the columns. Determine whether any of the columns hold missing values.
```{r}
cat("(a) Data structure:\n")
str(ais)
cat("\nMissing values per column:\n")
print(colSums(is.na(ais)))
```
(b) Make a table that shows the numbers of males and females for each different sport. In which sports is there a large imbalance (e.g., by a factor of more than 2:1) in the numbers of the two sexes?
```{r}
gender_table <- as.data.frame.matrix(table(ais$sport, ais$sex))
gender_table$Ratio <- round(gender_table$m / gender_table$f, 2)
imbalanced <- gender_table[gender_table$Ratio > 2 | gender_table$Ratio < 0.5, ]
cat("\n(b) Sports with gender imbalance (>2:1 or <1:2 ratio):\n")
print(imbalanced)
```
MB.Ch1.6.Create a data frame called Manitoba.lakes that contains the lake's elevation (in meters above sea level) and area (in square kilometers) as listed below. Assign the names of the lakes using the `row.names()` function.

```{=latex}
\begin{table}[h!]
  \begin{center}
\begin{tabular}{l|r|r} \hline
                & elevation & area \\ \hline
Winnipeg        &     217   & 24387 \\
Winnipegosis    &     254   & 5374 \\
Manitoba        &     248   & 4624 \\
SouthernIndian  &     254   & 2247 \\
Cedar           &     253   & 1353 \\
Island          &     227   & 1223 \\
Gods            &     178   & 1151 \\
Cross           &     207   & 755 \\
Playgreen       &    217    & 657 \\\hline
\end{tabular}
\end{center}
\end{table}
```

(a) Use the following code to plot `log2(area)` versus elevation, adding labeling information (there is an extreme value of area that makes a logarithmic scale pretty much essential):

```{r}
Manitoba.lakes <- data.frame(
  elevation = c(217, 254, 248, 254, 253, 227, 178, 207, 217),
  area = c(24387, 5374, 4624, 2247, 1353, 1223, 1151, 755, 657)
)
rownames(Manitoba.lakes) <- c("Winnipeg", "Winnipegosis", "Manitoba", 
                              "SouthernIndian", "Cedar", "Island", 
                              "Gods", "Cross", "Playgreen")
#attach(Manitoba.lakes)
#plot(log2(area) ~ elevation, pch=16, xlim=c(170,280))
## NB: Doubling the area increases log2(area) by 1.0
#text(log2(area) ~ elevation, labels=row.names(Manitoba.lakes), pos=4)
#text(log2(area) ~ elevation, labels=area, pos=2) 
#title("Manitoba's Largest Lakes")
plot(log2(area) ~ elevation, data = Manitoba.lakes, pch = 16, 
     xlim = c(170, 280), xlab = "Elevation (m)", ylab = "log2(Area)")
text(Manitoba.lakes$elevation, log2(Manitoba.lakes$area), 
     labels = rownames(Manitoba.lakes), pos = 4, cex = 0.8)
text(Manitoba.lakes$elevation, log2(Manitoba.lakes$area), 
     labels = Manitoba.lakes$area, pos = 2, cex = 0.8)
title("Manitoba Lakes (log2 scale)")

cat("Y-axis : Each unit increase in log2(area) represents a doubling of area\n\n")
cat("Labels : Lake names (right) link points to lakes; raw area values (left) show actual sizes, clarifying the log - scale compression of large area ranges.\n\n")
cat("This setup uses vertical distance to represent multiplicative (doubling) area changes, making extreme values (e.g., Winnipeg’s 24,387 km²) comparable to smaller lakes.\n\n")
```

Devise captions that explain the labeling on the points and on the y-axis. It will be necessary to explain how distances on the scale relate to changes in area.

(b) Repeat the plot and associated labeling, now plotting area versus elevation, but specifying `ylog=TRUE` in order to obtain a logarithmic y-scale.
```{r}
#plot(area ~ elevation, pch=16, xlim=c(170,280), ylog=T)
#text(area ~ elevation, labels=row.names(Manitoba.lakes), pos=4, ylog=T)
#text(area ~ elevation, labels=area, pos=2, ylog=T) 
#title("Manitoba's Largest Lakes")
plot(area ~ elevation, data = Manitoba.lakes, pch = 16, log = "y",
     xlim = c(170, 280), ylim = c(500, 30000),
     xlab = "Elevation (m)", ylab = "Area (km², log scale)")
text(Manitoba.lakes$elevation, Manitoba.lakes$area, 
     labels = rownames(Manitoba.lakes), pos = 4, cex = 0.8)
title("Manitoba Lakes (logarithmic y-scale)")
```

MB.Ch1.7. Look up the help page for the R function `dotchart()`. Use this function to display the areas of the Manitoba lakes (a) on a linear scale, and (b) on a logarithmic scale. Add, in each case, suitable labeling information.
```{r}
#dotchart(log2(area))
dotchart(Manitoba.lakes$area, labels = rownames(Manitoba.lakes),
         main = "Lake Areas (Linear Scale)",
         xlab = "Area (km²)", pch = 16)

```

MB.Ch1.8. Using the `sum()` function, obtain a lower bound for the area of Manitoba covered by water.
```{r}
total_water_area <- sum(Manitoba.lakes$area)
cat("Lower bound for Manitoba water area:", 
    format(total_water_area, big.mark = ","), "km²\n")
cat("(Sum of the 9 largest lakes)")
```